Experience Report: Distributed Computing for the Common Man

### 2010-03-18 doc/icfp-2010/main.tex
*** empty log message ***

The distributing computing library presented in this paper, is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of APIs that allows
execution of large-scale parallelizable computations, very relevant to
our application needs, (and presumably relevant to a variety of
real-world applications) over multiple cores in the same machine or
over a network of machines. It relieves the user from implementing
details of fault-tolerance, task distribution and parallelization. The
most striking feature of the library is its generality. The user needs
to change no more than a couple of lines in his code when he moves his
large-scale computation from a multi-core machine to a network of
machines. Further, the library provides exciting flexibilities as it
supports polymorphic functions, and rich features such as
fault-tolerance that become a cause for concern in any distributed
programming environment. It is worth noting that the library is not
targeted at applications running on server farms, crunching enormous
amounts of data. Neither it is limited to research labs possessing a
few number of powerful computing infrastructure. It serves a variety
of users and a wide spectrum of needs, from desktop PCs to network of
machines, and hence we find title \emph{Distributed Computing for the
  Common Man}, apt and appropriate.

@@

\emph{Distributed Computing:} 
A typical distributed computing library, in its minimal implementation
should provide the following:
\begin{itemize}
\item A notion of a \of{task} which denotes the computation to be
  performed in a distributed manner
\item A set of machines or programs called \of{workers} that perform
  the tasks, producing results
\item A machine or a program called a \of{master} that is incharge
  of distributing the tasks among the workers and managing results
  produced by the workers
\end{itemize}
In addition to the above, distributed computing environments also
implement mechanisms for fault-tolerance and efficient storage and
distribution of tasks. This is required to handle network failures
that may occur, as well as to optimize the usage of machines in the
network. Another concern of importance is the transmission of messages
over the network. This requires efficient serialization (or
marshalling) of data that can be reliably reconstructed when
transmitted over different computing environments.  It is desirable to
maintain architecture independence while transmitting marshalled data,
as machines in a distributed computing environment more often run on
different hardware architectures and make use of varying software
platforms. For example, a computer running on a hardware architecture
should be able to reliably reconstruct the data irrespective of
endianness. We carefully exploit the marshalling provided by \Ocaml\,
in particular the restrictions involved in marshalling functions, in
order to achieve maximally reliable transmission of messages over a
network.

Google's approach was also one of our inspirations to design and
implement a generic distributed computing library for functional
programming. Ironically, Google's approach itself was inspired by
functional programming primitives.

The paper is organized as follows. In Section~\ref{sec:API}, we
describe the main idea behind the generic API, and introduce
\of{master} and \of{worker} programs. In Section~\ref{sec:derived}, we
derive more interesting functions from the generic
API. Section~\ref{sec:implem} delves into the implementation details
of our library, while Section~\ref{sec:experiments} illustrates the
potential of the presented library through experimental evaluation. We
draw some conclusions in Section~\ref{sec:future} and outline our
future work.

% master and worker terminology introduced here
% a few technical details about marshaling should be explained

@@

The main idea is to start with a generic function, with the
following interface: 
% TODO: rename this function?
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
% TODO be even more explicit about the introduction of the word `task'
This function takes two functions, \ocaml{worker} and \ocaml{master}, and a
list of \emph{tasks} of type \ocaml{('a * 'c) list}.
The function \ocaml{worker} is applied to the first component for each task.
Whenever it completes, the task and the result are passed to the
function \ocaml{master}. This may possibly generate a new list of
tasks. The process ends when all computations are done.

Function \ocaml{worker} should be pure and is executed in parallel.
Function \ocaml{master}, on the contrary, can be impure and is never
executed in parallel with other \ocaml{master}s. Typically,
\ocaml{master} stores and processes information locally.
The next section will describe the usage of this generic interface to
perform traditional map and fold operations.

Actually, our library provides more than just a single \ocaml{compute}
function as above, but instead provides five different versions
depending on the execution context. The five possible contexts are the
following: 
\begin{enumerate}
\item \textbf{purely sequential execution:}
  this is intended for validating the results;

@@

\item \textbf{master and workers are different programs, compiled with
    the same version of \Ocaml:} 
  in this case, we can no longer marshal closures but we can still
  marshal polymorphic values. As a consequence, 
  the \ocaml{compute} function is split into two
  polymorphic functions, to implement the master and workers respectively:
  \begin{ocaml}
val master : 
  ('a * 'c -> 'b -> ('a * 'c) list) -> 
  ('a * 'c) list -> unit
val worker : ('a -> 'b) -> unit -> unit
  \end{ocaml}

\item \textbf{master and workers are different programs, not even
    compiled with the same version of \Ocaml:} in this case, we can no
  longer use marshaling, so the
  \ocaml{compute} function is split into two monomorphic functions
  over strings: 
  \begin{ocaml}
val master : 
  (string * 'c -> string -> (string * 'c) list) -> 
  (string * 'c) list -> unit
val worker : (string -> string) -> unit -> unit
  \end{ocaml}
  Note that the second component of each task is still polymorphic (of
  type \ocaml{'c} here).
\end{enumerate}

Our library is organized into three modules: \of{Sequential} for the
pure sequential implementation, \of{Cores} for multiple cores on the
same machine and \of{Network} for a network of machines, respectively.
The \of{Network} module itself is organized into three sub-modules,
corresponding to contexts 3, 4 and 5 above. The different
API functions are summarized in the following table (their signatures
being given above):
\begin{center}
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    \ocaml{Sequential} & \ocaml{Cores} &
    \multicolumn{3}{|c|}{\ocaml{Network}} 
    \\\cline{3-5}
    &       & \ocaml{Same} & \ocaml{Poly} & \ocaml{Mono} \\\hline
    \ocaml{compute}  & \ocaml{compute}   & \ocaml{compute}  &
    \ocaml{master}  & \ocaml{master}  \\
    & & & \ocaml{worker}  & \ocaml{worker} \\\hline
  \end{tabular}
\end{center}

@@

In most cases, the easiest way to parallelize execution it to make use
of map and fold operations over lists, where processing of the list
elements are done in parallel.  Using the generic API described in the
previous section to do that is not very convenient.
Hence we provide the most commonly used list operations in our API.

The most obvious operation is the traditional map operation over
lists, that is:
\begin{ocaml}
  val map : ('a -> 'b) -> 'a list -> 'b list
\end{ocaml}
The next natural operation is a combination of map and fold
operations, that is a function like
\begin{ocaml}
 val map_fold :
   map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
   'c -> 'a list -> 'c
\end{ocaml}
which, given two functions, an accumulator \of{a} and a list \of{l}, computes
\begin{equation}\label{eq:map-fold}
  \of{fold} ... (\of{fold} (\of{fold} ~ a ~ (\of{map} ~ x_1)) (\of{map} ~ x_2))
  ... (\of{map} ~ x_n)
\end{equation}
for some permutation $[x_1,x_2,...,x_n]$ of the list \of{l}.

We assume that the \of{map} operations are always performed in parallel.
Regarding \of{fold} operations, we distinguish two cases:
\begin{itemize}
\item either \of{fold} operations are computationally less expensive
  than \of{map} and we perform them locally;
\item or \of{fold} operations are computationally expensive and we
  perform them in parallel.
\end{itemize}
We thus provide two functions \of{map_local_fold} and \of{map_remote_fold}.

@@

% LocalWords:  parallelize functor parameterized indices endianness monomorphic

>>>>

The distributing computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of APIs that allows
execution of large-scale parallelizable computations, very relevant to
our application needs (and presumably relevant to a variety of
real-world applications), over multiple cores in the same machine or
over a network of machines. The most important features of our library
are the following:
\begin{itemize}
\item \emph{genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{simplicity}: 
  switching between multiple cores and a network of
  machine is as simple as changing a couple of lines of code;
\item \emph{task distribution and fault-tolerance}: 
  it relieves the user from implementing task distribution routines
  and reliable fault-tolerance mechanism.
\end{itemize}
It is worth noting that the library is not
targeted at applications running on server farms, crunching enormous
amounts of data. Neither it is limited to research labs possessing a
few number of powerful computing infrastructure. It serves a variety
of users and a wide spectrum of needs, from desktop PCs to network of
machines, and hence the title.

@@

\paragraph{Distributed Computing.}
A typical distributed computing library
provides the following (borrowing some terminology from Google's
Mapreduce~\cite{mapreduce}):
\begin{itemize}
\item A notion of a \emph{task} which denotes the atomic computation to be
  performed in a distributed manner;
\item A set of machines or programs called \emph{workers} that perform
  the tasks, producing results;
\item A single program called a \emph{master} that is in charge
  of distributing the tasks among the workers and managing results
  produced by the workers.
\end{itemize}
In addition to the above, distributed computing environments also
implement mechanisms for fault-tolerance, efficient storage and
distribution of tasks. This is required to handle network failures
that may occur, as well as to optimize the usage of machines in the
network. Another concern of importance is the transmission of messages
over the network. This requires efficient serialization (or
marshalling) of data that can be reliably reconstructed when
transmitted over different computing environments.  It is desirable to
maintain architecture independence while transmitting marshalled data,
as machines in a distributed computing environment often run on
different hardware architectures and make use of varying software
platforms. For example, endianness may be different across machines on
the network.

\paragraph{A Functional Programming Approach.}
Our work was initially inspired by Google's
Mapreduce\footnote{Ironically, Google's approach itself was inspired
  by functional programming primitives.}. However, our functional
programming environment allows us to be more generic. 
The main idea behind our approach is that
workers may implement any polymorphic function
\begin{ocaml}
  worker: 'a -> 'b
\end{ocaml}
where \of{'a} denotes the type of tasks and \of{'b} the type of results.
Then the master is given a list of initial tasks, together with a
function to handle the results:
\begin{ocaml}
  master: ('a -> 'b -> 'a list) -> 'a list -> unit
\end{ocaml}
The function passed to the master is applied whenever a result is
available and may in turn generate new tasks (hence the return type
\of{'a list}).  The master is executed as long as there are incomplete
tasks.

Our library makes use of \Ocaml's marshaling capabilities, as much as
possible. Whenever master and worker executables are exactly the same,
we can marshal polymorphic values and closures. However, it is not
always possible to have master and workers running the same
executable. In that case, we cannot marshal closures anymore but we
can still marshal polymorphic values as long as the same version of
\Ocaml\ is used to compile master and workers. When different versions
of \Ocaml\ are used, we can no longer marshal values but we can still
transmit strings between master and workers. Our library adapts to all
these situations, by providing several APIs.

The paper is organized as follows. In Section~\ref{sec:API}, we
describe the main idea behind the generic API, and introduce
\of{master} and \of{worker} programs. In Section~\ref{sec:derived}, we
derive more interesting functions from the generic
API. Section~\ref{sec:implem} delves into the implementation details
of our library, while Section~\ref{sec:experiments} illustrates the
potential of the presented library through experimental evaluation. We
compare our approach with relevant related work in
Section~\ref{sec:future} and outline our future work.

@@

The main function in our API follows the idea sketched in the
introduction. It has the following signature:
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
Tasks are pairs, of type \of{'a * 'c}, where the first component is
passed to the worker and the second component is local to the master.
Function \ocaml{worker} should be pure and is executed in parallel.
Function \ocaml{master}, on the contrary, can be impure and is never
executed in parallel with other \ocaml{master}s. Function
\ocaml{master} always processes information locally, typically storing
it to internal data structures.
The next section will describe the usage of this generic interface to
perform traditional map and fold operations.

Actually, our library provides not just a single \ocaml{compute}
function as above, but instead five different versions
depending on the execution context. 
The five possible contexts are the following: 
\begin{enumerate}
\item \textbf{purely sequential execution:}
  this is intended for validating the results;

@@

\item \textbf{master and workers are different programs, compiled with
    the same version of \Ocaml:} 
  we can no longer marshal closures but we can still
  marshal polymorphic values. As a consequence, 
  the \ocaml{compute} function is split into two
  polymorphic functions, to implement the master and workers 
  respectively:%
\begin{ocaml}
val master : 
  ('a * 'c -> 'b -> ('a * 'c) list) -> 
  ('a * 'c) list -> unit
val worker : ('a -> 'b) -> unit -> unit
\end{ocaml}
\item \textbf{master and workers are different programs, not even
    compiled with the same version of \Ocaml:} we can no
  longer use marshaling, so the
  \ocaml{compute} function is split into two monomorphic functions
  over strings:%
\begin{ocaml}
val master : 
  (string * 'c -> string -> (string * 'c) list) -> 
  (string * 'c) list -> unit
val worker : (string -> string) -> unit -> unit
\end{ocaml}
  Note that the second component of each task is still polymorphic (of
  type \ocaml{'c} here), since it is local to the master.
\end{enumerate}

Our library is organized into three modules: \of{Sequential} for the
pure sequential implementation, \of{Cores} for multiple cores on the
same machine and \of{Network} for a network of machines, respectively.
The \of{Network} module itself is organized into three sub-modules,
corresponding to contexts 3, 4 and 5 above. The different
API functions are summarized in the following table (their signatures
being given above):
\begin{center}
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    \ocaml{Sequential} & \ocaml{Cores} &
    \multicolumn{3}{|c|}{\ocaml{Network}} 
    \\\cline{3-5}
    &       & \ocaml{Same} & \ocaml{Poly} & \ocaml{Mono} \\\hline
    \ocaml{compute}  & \ocaml{compute}   & \ocaml{compute}  &
    \ocaml{master}  & \ocaml{master}  \\
    & & & \ocaml{worker}  & \ocaml{worker} \\\hline
  \end{tabular}
\end{center}
The following section introduces other interesting APIs that can be
derived from this generic API.

@@

In most cases, the easiest way to parallelize execution it to make use
of map and fold operations over lists, where processing of the list
elements are done in parallel.  To facilitate such a processing, we
now derive the most commonly used list operations from our generic
API.

The most obvious operation is the traditional map operation over
lists, that is:
\begin{ocaml}
  val map : ('a -> 'b) -> 'a list -> 'b list
\end{ocaml}
The next natural operation is a combination of map and fold
operations, that is a function like
\begin{ocaml}
 val map_fold :
   map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
   'c -> 'a list -> 'c
\end{ocaml}
which, given two functions, an accumulator \of{a} and a list \of{l}, computes
\begin{equation}\label{eq:map-fold}
  \of{fold} ... (\of{fold} (\of{fold} ~ a ~ (\of{map} ~ x_1)) (\of{map} ~ x_2))
  ... (\of{map} ~ x_n)
\end{equation}
for some permutation $[x_1,x_2,...,x_n]$ of the list \of{l}.
We assume that the \of{map} operations are always performed in parallel.
Regarding \of{fold} operations, we distinguish two cases:
\begin{itemize}
\item either \of{fold} operations are computationally less expensive
  than \of{map} and we perform them locally;
\item or \of{fold} operations are computationally expensive and we
  perform them in parallel.
\end{itemize}
We thus provide two functions \of{map_local_fold} and \of{map_remote_fold}.

@@

% LocalWords:  parallelize functor parameterized indices endianness monomorphic
% LocalWords:  genericity executables
